Tissue Enrichment Analysis: TEA

### 2016-03-21 tests/hgf_benchmarking/doc/tea_main_text.tex
More writing.

\begin{abstract} % abstract
\parttitle{Background} %if any
Over the last ten years, there has been an explosive development in tools
capable of measuring gene expression. These tools generate a large number
of gene targets, but understanding these datasets, and forming hypotheses
based on them remains challenging. One way to analyze these datasets is to
associate ontologies, which are controlled, hierarchical, descriptive vocabularies
with genes and to look for enrichment of specific terms. Although gene ontology
(GO) is available for   \emph{C. elegans}, this ontology does not include
anatomy or physiology information.
\parttitle{Results} %if any
We have developed an enrichment analysis tool for the   \emph{C. elegans} tissue
ontology, is available on the web via WormBase and is available for download 
using Python's standard pip installer. In order to cut down on verbosity, we
have come up with three straightforward filtering criteria that slim the ontology
by almost tenfold. 
\parttitle{Conclusions} %if any
Our Tissue Enrichment Analysis (TEA), which can be found at www.WormBase.org/tea, 
uses a standard hypergeometric function to test a slimmed-down   \emph{C. elegans} tissue 
ontology and provides users with a text and graphic representation of the results. 
\end{abstract}

@@

	Once the probability of drawing the labels has been quantified, we apply a standard FDR correction using a Benjamini-Hochberg step-up algorithm \cite{Benjamini1995}. Genes that have a q-value less than a given alpha are considered significant. Our default setting is to set the alpha threshold at 0.1, but users will be able to modify this value either in batch or in our web application. The program returns a text-based table showing the tissues that tested significant, along with their associated q-value, the expected number of hits for a list of that size, the observed number of hits and the enrichment fold change (observed hits / expected hits). Finally, the program can also return a bar chart of the enrichment fold change for the fifteen tissues with the lowest measured q-values.
	Our software is implemented in an easy to use GUI within WormBase. Users input a gene-list using any valid gene name for   \emph{C. elegans}. These names are processed into standard WBIDs and the result is displayed in the same window in an easy to read format containing all the relevant information, and a graph of the results is also displayed (see Figure ~\ref{fig:GUIresults}).

\subsection*{Validation of the algorithm and parameter selection}
	We wanted to select a dictionary which included enough terms to be specific beyond the largest  \emph{C. elegans}tissues, yet would minimize the number of spurious results and which had a good dynamic range in terms of enrichment fold-change. Selection of a dictionary based only on minimization of spurious results would result in a dictionary with a large number of annotations per tissue, and would therefore include only the major tissues. On the other hand, selecting a dictionary that can detect smaller tissues will bias us towards tissues with lesser annotations. To our knowledge there is no good method for assessing false-positive or false-negative results for annotations.To help us select an appropriate dictionary and validate our tool, we found a set of 30 gold standards based on microarray and RNA-seq literature which are believed to be enriched in specific tissues \cite{Gaudet2004a, Spencer2011, Cinar2005, Watson2008a, Pauli2006, Portman2004, Fox2007, Smith2010}. Some of these studies have been used to annotate gene expression, and so they did not constitute an independent testing set. To correct this flaw, we built a clean dictionary that specifically excluded all annotation evidence that came from these studies.

@@

\subsection*{A brief example}
We applied our tool to the RNA-seq datasets developed by Engelmann et al.  \cite{Engelmann2011} to gain further understanding of the biology underlying these datasets. Engelmann et al.  exposed young adult worms to 5 different pathogenic bacteria or fungi for 24 hours, after which mRNA was extracted from the worms for sequencing. We ran TEA on the genes Engelmann \emph{et al} identified as up- or down-regulated. Initially we noticed that genes that are down-regulated tend to be twice better annotated on average than genes that were up-regulated, suggesting that our understanding of the worm immune system is scarce, in spite of important advances made over the last decade. Strikingly, three out of the five samples showed enrichment of neuronal tissues or neuronal precursor tissues amongst the down-regulated genes (see \ref{fig:Dcon, fig:Hsp, fig:Smar}). A possible explanation for this might be that the infected worms are sick and the neurons are beginning to shut down; an alternative hypothesis would be that the worm is down-regulating specific neuronal pathways as a behavioral response against the pathogen. Indeed, several studies \cite{Meisel2014, Zhang2005} have provided evidence that   \emph{C. elegans} uses chemosensory neurons to identify pathogens. Interestingly, one bacterium did not exhibit the same pattern of down-regulation of neuronal-associated genes. \emph{E. faecalis} showed increased expression of genes associated with neuronal tissues, hinting that \emph{E. faecalis} may have a different pathogenic profile. Up-regulated tissues, when detected, almost always included the hypodermis and excretory duct.
	Our results highlight the involvement of various   \emph{C. elegans} neuronal tissues in pathogen defense and/or illness.

@@

\section*{Methods}
\subsection*{Filtering nodes}
\subsubsection*{Defining a Similarity Metric}
In order to identify redundant sisters, we defined the following similarity metric:
\begin{eqnarray}\label{similarity def}
	s_i = \frac{|g_i|}{|\bigcup_{i= 0}^k g_i|}
\end{eqnarray}
Where $s_i$ is the similarity for a tissue $i$ with $k$ sisters; $g_i$ refers to the set of tissues associated with tissue $i$ and $|g|$ refers to the cardinality of set $g$. For a given set of sisters, we called them redundant if they exceeded a given similarity threshold. We envisioned two possible criteria and built different dictionaries using each one. Under a threshold criteron `any' with parameter $S$ between $(0, 1)$, a given set of sisters $j$ was considered redundant if the condition
\begin{eqnarray}\label{any threshold}
	s_{i, j} > S
\end{eqnarray}
was true for any sister $i$ in set $j$. Under a threshold criterion `avg' with parameter $S$, a given set of sisters $j$ was considered redundant if the condition
\begin{eqnarray}\label{avg threshold}
	\mathrm{E}[s_i]_j > S
\end{eqnarray}
was true for the set of sisters $j$ (see Figure~\ref{fig:simdiagram}).

@@

  \captionsetup{width= 0.95\textwidth}
  \caption{
  	\csentence{Schematic representation of trimming filters for an acyclical ontology.}
    \csentence{\textbf{a)}} The parent node (green) contains at least as many annotations as the union of the two sisters. These two sisters share annotations extensively. Therefore they are too similar and should be removed.
	\csentence{\textbf{b)}} Nodes with less than a threshold number of genes are trimmed (light red) and discarded from the dictionary. Here, the example threshold is 25 genes.
	\csentence{\textbf{c)}} We trim parent nodes recursively, starting from the root, if all their daughter nodes have more than the threshold number of annotations (light red).
  }
\end{figure}

@@

\begin{figure}
	\centering
    \includegraphics[width=0.95\textwidth]{figures/guiresults.png}
  	\captionsetup{width= 0.95\textwidth}
 	\caption{\csentence{Screenshot of results from the web GUI.}
  After inputting a gene-list, the user is provided with the results. 
  An HTML table is output with hyperlinks to the ontology terms. A 
  publication-ready graph is provided below, which can be saved by 
  dragging to the desktop. Finally, lists of the genes used and discarded
  for the analysis are also presented.
  }
  \label{fig:GUIresults}
\end{figure}

@@

%pan neuronal
\begin{figure}
  \begin{subfigure}[b]{0.95\textwidth}
    \includegraphics[width=0.95\textwidth]{WBPaper00031532_Larva_Pan_Neuronal_Enriched_WBbt_0003679_1603_33cutoff.png}
    \caption{}
    \label{fig:PanNeuronalWatson33}
  \end{subfigure}
  %
  \begin{subfigure}[b]{0.95\textwidth}
    \includegraphics[width=0.95\textwidth]{WBPaper00031532_Larva_Pan_Neuronal_Enriched_WBbt_0003679_1603_50cutoff}
    \caption{}
    \label{fig:PanNeuronalWatson50}
  \end{subfigure}
  \captionsetup{width= 0.95\textwidth}
  \caption{\csentence{Comparison of Enrichment Results for two dictionaries for a pan-neuronal enriched gene set from ~\cite{Watson2008a}.}
   \textbf{a)} Dictionary with cutoff: 33; threshold: 0.95; method: `any'
   \textbf{b)} Dictionary with cutoff: 50; threshold: 0.95; method: `any'
   }
   \label{fig:interagree}
\end{figure}

%agreements
\begin{figure}
  \begin{subfigure}[b]{0.95\textwidth}
    \includegraphics[width=0.95\textwidth]{WBPaper00024970_GABAergic_neuron_specific_WBbt_0005190_247_33cutoff.png}
    \caption{}
    \label{fig:agreement1}
  \end{subfigure}
  %
  \begin{subfigure}[b]{0.95\textwidth} 	\includegraphics[width=0.95\textwidth]{WBPaper00037950_GABAergic-motor-neurons_larva_enriched_WBbt_0005190_132_33cutoff.png}
    \caption{}
    \label{fig:agreement2}
  \end{subfigure}
  \captionsetup{width= 0.95\textwidth}
  \caption{\csentence{Independently derived genesets show similar results within a given dictionary}
  \textbf{a)} Gene set from~\cite{Watson2008a}. Dictionary with cutoff: 33; threshold: 0.95; method: `any'
  \textbf{b)} Gene set from~\cite{Spencer2011}. Dictionary with cutoff: 50; threshold: 0.95; method: `any'
   }
   \label{fig:intragree}
\end{figure}

@@

%% Use of \listoftables is discouraged.
%%
\section*{Tables}
\begin{table}[h!]
\caption{Parameter specifications and number of tissues for all dictionaries.}
	\csvautotabular{figures/TissueNumbers.csv}
	\label{tab:DictionarySpecs}
\end{table}

>>>>

\begin{abstract} % abstract
\parttitle{Background} %if any
Over the last ten years, there has been an explosive development in tools
capable of measuring gene expression. These tools generate a large number
of gene targets, but understanding these datasets, and forming hypotheses
based on them remains challenging. One way to analyze these datasets is to
associate ontologies, which are controlled, hierarchical, descriptive vocabularies
with genes and to look for enrichment of specific terms. Although gene ontology
(GO) is available for   \emph{Caenorhabditis elegans}, this ontology does not include
anatomy or physiology information.
\parttitle{Results} %if any
We have developed an enrichment analysis tool for the   \emph{C. elegans} tissue
ontology, is available on the web via WormBase and is available for download 
using Python's standard pip installer. In order to cut down on verbosity, we
have come up with three straightforward filtering criteria that slim the ontology
by almost tenfold. 
\parttitle{Conclusions} %if any
Our Tissue Enrichment Analysis (TEA), which can be found at www.WormBase.org/tea, 
uses a standard hypergeometric function to test a slimmed-down   \emph{C. elegans} tissue 
ontology and provides users with a text and graphic representation of the results. 
\end{abstract}

@@

	Once the probability of drawing the labels has been quantified, we apply a standard FDR correction using a Benjamini-Hochberg step-up algorithm \cite{Benjamini1995}. Genes that have a q-value less than a given alpha are considered significant. Our default setting is to set the alpha threshold at 0.1, but users will be able to modify this value either in batch or in our web application. The program returns a text-based table showing the tissues that tested significant, along with their associated q-value, the expected number of hits for a list of that size, the observed number of hits and the enrichment fold change (observed hits / expected hits). Finally, the program can also return a bar chart of the enrichment fold change for the fifteen tissues with the lowest measured q-values.
	Our software is implemented in an easy to use GUI within WormBase. Users input a gene-list using any valid gene name for  \emph{C. elegans}. These names are processed into standard WBIDs and the result is displayed in the same window in an easy to read format containing all the relevant information, and a graph of the results is also displayed (see Figure ~\ref{fig:GUIresults}). Anatomy terms are displayed in human-readable format followed by their unique WBbt ID.

\subsection*{Validation of the algorithm and parameter selection}
	We wanted to select a dictionary which included enough terms to be specific beyond the largest  \emph{C. elegans}tissues, yet would minimize the number of spurious results and which had a good dynamic range in terms of enrichment fold-change. Selection of a dictionary based only on minimization of spurious results would result in a dictionary with a large number of annotations per tissue, and would therefore include only the major tissues. On the other hand, selecting a dictionary that can detect smaller tissues will bias us towards tissues with lesser annotations. To our knowledge there is no good method for assessing false-positive or false-negative results for annotations.
	% Gold-standard data sets were annotated gene lists derived from the corresponding Expression Cluster data available in the WS252 version of WormBase.
	To help us select an appropriate dictionary and validate our tool, we used a set of 30 gold standards based on microarray and RNA-seq literature which are believed to be enriched in specific tissues \cite{Gaudet2004a, Spencer2011, Cinar2005, Watson2008a, Pauli2006, Portman2004, Fox2007, Smith2010}. These data sets are annotated gene lists derived from the corresponding Expression Cluster data available in the WS252 version of WormBase. Some of these studies have been used to annotate gene expression, and so they did not constitute an independent testing set. To correct this flaw, we built a clean dictionary that specifically excluded all annotation evidence that came from these studies.

@@

\subsection*{A brief example}
We applied our tool to the RNA-seq datasets developed by Engelmann et al.  \cite{Engelmann2011} to gain further understanding of the biology underlying these datasets. Engelmann et al.  exposed young adult worms to 5 different pathogenic bacteria or fungi for 24 hours, after which mRNA was extracted from the worms for sequencing. We ran TEA on the genes Engelmann \emph{et al} identified as up- or down-regulated. Initially we noticed that genes that are down-regulated tend to be twice better annotated on average than genes that were up-regulated, suggesting that our understanding of the worm immune system is scarce, in spite of important advances made over the last decade. Strikingly, three out of the five samples showed enrichment of neuronal tissues or neuronal precursor tissues amongst the down-regulated genes (see \ref{fig:Dcon}, \ref{fig:Hsp},\ref{fig:Smar}). A possible explanation for this might be that the infected worms are sick and the neurons are beginning to shut down; an alternative hypothesis would be that the worm is down-regulating specific neuronal pathways as a behavioral response against the pathogen. Indeed, several studies \cite{Meisel2014, Zhang2005} have provided evidence that   \emph{C. elegans} uses chemosensory neurons to identify pathogens. Interestingly, one bacterium did not exhibit the same pattern of down-regulation of neuronal-associated genes. \emph{E. faecalis} showed increased expression of genes associated with neuronal tissues, hinting that \emph{E. faecalis} may have a different pathogenic profile. Up-regulated tissues, when detected, almost always included the hypodermis and excretory duct.
	Our results highlight the involvement of various   \emph{C. elegans} neuronal tissues in pathogen defense and/or illness.

@@

\section*{Methods}
\subsection*{Fetching annotation terms}
We used the WormBase gene expression data, which includes annotated descriptions of spatial-temporal expression patterns of genes, to build our dictionary. Gene lists per anatomy term were extracted from a solr document store of gene expression data from the WS252 database provided by WormBase (PMID: 26578572). We used the solr document store because it provided a convenient access to expression data that included inferred annotations. That is, for each anatomy term, the expression gene list includes genes that were directly annotated to the term, as well as those that were annotated to the term's descendant terms (if there were any). Descendant terms were those connected with the focus term by is\_a/part\_of relationship chains defined in the anatomy term ontology hierarchy.
\subsection*{Filtering nodes}
\subsubsection*{Defining a Similarity Metric}
In order to identify redundant sisters, we defined the following similarity metric:
\begin{eqnarray}\label{similarity def}
	s_i = \frac{|g_i|}{|\bigcup_{i= 0}^k g_i|}
\end{eqnarray}
Where $s_i$ is the similarity for a tissue $i$ with $k$ sisters; $g_i$ refers to the set of tissues associated with tissue $i$ and $|g|$ refers to the cardinality of set $g$. For a given set of sisters, we called them redundant if they exceeded a given similarity threshold. We envisioned two possible criteria and built different dictionaries using each one. Under a threshold criteron `any' with parameter $S$ between $(0, 1)$, a given set of sisters $j$ was considered redundant if the condition
\begin{eqnarray}\label{any threshold}
	s_{i, j} > S
\end{eqnarray}
was true for any sister $i$ in set $j$. Under a threshold criterion `avg' with parameter $S$, a given set of sisters $j$ was considered redundant if the condition
\begin{eqnarray}\label{avg threshold}
	\mathrm{E}[s_i]_j > S
\end{eqnarray}
was true for the set of sisters $j$ (see Figure~\ref{fig:simdiagram}).

@@

  \captionsetup{width= 0.95\textwidth}
  \caption{
  	\csentence{Schematic representation of trimming filters for an acyclical ontology.}
    \csentence{\textbf{a)}} The parent node (green) contains at least as many annotations as the union of the two sisters. These two sisters share annotations extensively. Therefore they are too similar and should be removed.
	\csentence{\textbf{b)}} Nodes with less than a threshold number of genes are trimmed (red) and discarded from the dictionary. Here, the example threshold is 25 genes. Nodes $\epsilon, \zeta, \eta$, shown in red will be removed.
	\csentence{\textbf{c)}} We trim parent nodes recursively, starting from the root, if all their daughter nodes have more than the threshold number of annotations. Nodes in grey ($\epsilon, \zeta, \eta$) were removed in the previous step. Nodes $\alpha, \beta$ shown in red will be trimmed because each one has a complete daughter set. Only nodes $\gamma$ and $\delta$ will be used to generate the static dictionary. 
  }
\end{figure}

@@

\begin{figure}
	\centering
    \includegraphics[width=0.95\textwidth]{figures/guiresults.png}
  	\captionsetup{width= 0.95\textwidth}
 	\caption{\csentence{Screenshot of results from the web GUI.}
  After inputting a gene-list, the user is provided with the results. 
  An HTML table is output with hyperlinks to the ontology terms. A 
  publication-ready graph is provided below, which can be saved by 
  dragging to the desktop. The graph and the table show anatomy terms in
  human-readable format, followed by their unique WBbt ID. Finally, lists 
  of the genes used and discarded for the analysis are also presented.
  }
  \label{fig:GUIresults}
\end{figure}

@@

%pan neuronal
\begin{figure}
  \begin{subfigure}[b]{0.95\textwidth}
    \includegraphics[width=0.95\textwidth]{WBPaper00031532_Larva_Pan_Neuronal_Enriched_WBbt_0003679_1603_33cutoff.png}
    \caption{}
    \label{fig:PanNeuronalWatson33}
  \end{subfigure}
  %
  \begin{subfigure}[b]{0.95\textwidth}
    \includegraphics[width=0.95\textwidth]{WBPaper00031532_Larva_Pan_Neuronal_Enriched_WBbt_0003679_1603_50cutoff}
    \caption{}
    \label{fig:PanNeuronalWatson50}
  \end{subfigure}
  \captionsetup{width= 0.95\textwidth}
  \caption{\csentence{Comparison of Enrichment Results for two dictionaries for a pan-neuronal enriched gene set from Watson, 2008~\cite{Watson2008a}.}
	Bar labels are anatomy terms in human-readable format, followed by their unique WBbt ID.
   \textbf{a)} Dictionary with cutoff: 33; threshold: 0.95; method: `any'
   \textbf{b)} Dictionary with cutoff: 50; threshold: 0.95; method: `any'
   }
   \label{fig:interagree}
\end{figure}

%agreements
\begin{figure}
  \begin{subfigure}[b]{0.95\textwidth}
    \includegraphics[width=0.95\textwidth]{WBPaper00024970_GABAergic_neuron_specific_WBbt_0005190_247_33cutoff.png}
    \caption{}
    \label{fig:agreement1}
  \end{subfigure}
  %
  \begin{subfigure}[b]{0.95\textwidth} 	\includegraphics[width=0.95\textwidth]{WBPaper00037950_GABAergic-motor-neurons_larva_enriched_WBbt_0005190_132_33cutoff.png}
    \caption{}
    \label{fig:agreement2}
  \end{subfigure}
  \captionsetup{width= 0.95\textwidth}
  \caption{\csentence{Independently derived genesets show similar results within a given dictionary}
  Bar labels are anatomy terms in human-readable format, followed by their unique WBbt ID.
  \textbf{a)} Gene set from Watson, 2008~\cite{Watson2008a}. Dictionary with cutoff: 33; threshold: 0.95; method: `any'
  \textbf{b)} Gene set from Spencer, 2011~\cite{Spencer2011}. Dictionary with cutoff: 50; threshold: 0.95; method: `any'
   }
   \label{fig:intragree}
\end{figure}

@@

%% Use of \listoftables is discouraged.
%%
\section*{Tables}
\begin{table}[h!]
\caption{Parameter specifications and number of tissues for all dictionaries.
The `Method' column refers to the trimming criterion for the similarity metric. We 
used two such criteria, `any' and `avg'.
`any': For a given sister set, if any sister had a similarity exceeding the corresponding
threshold, all sisters were removed from the final dictionary.
`avg': For a given sister set, if the average similarity across all the sisters in the set
was greater than the corresponding threshold, all sisters were removed from the final dictionary.}
	\csvautotabular{figures/TissueNumbers.csv}
	\label{tab:DictionarySpecs}
\end{table}
